# Pruning regression trees with `rpart`

```{r setup}
#| include: false
source("_common.R")
```



_**Important note**: As discussed in class, the K-fold CV methodology
implemented in the package `rpart` seems to consider 
a sequence of trees (or, equivalently, of complexity parameters)
based on the full training set. For more details
refer to the corresponding documentation: pages 12 and ff of the
package vignette, which can be accessed from `R` using the
command `vignette('longintro', package='rpart')`. 
For an alternative implementation of CV-based pruning, 
please see also the Section **"Pruning regression trees with `tree`"** below._ 

The stopping criteria generally used when fitting regression trees do not
take into account explicitly the complexity of the tree. Hence, we 
may end up with either an overfitting tree, or a very simple one, 
which typically results in a decline in the quality of the corresponding predictions. 
As discussed in class, one solution is to purposedly grow / train a very large overfitting
tree, and then prune it. One can also estimate the corresponding MSPE
of each tree in the prunning sequence and choose an optimal one. 
The function `rpart` implements this approach, and we illustrate it
below. 

We force `rpart` to build a very large tree via the arguments
of the function `rpart.control`. At the same time, to obtain a good
picture of the evolution of MSPE for different subtrees, we set the smallest 
complexity parameter to be considered by the cross-validation
experiment to a very low value (here we use `1e-8`).
```{r prune, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(rpart)
data(Boston, package = "MASS")
# split data into a training and
# a test set
set.seed(123456)
n <- nrow(Boston)
ii <- sample(n, floor(n / 4))
dat.te <- Boston[ii, ]
dat.tr <- Boston[-ii, ]

myc <- rpart.control(minsplit = 2, cp = 1e-5, xval = 10)
set.seed(123456)
bos.to <- rpart(medv ~ .,
  data = dat.tr, method = "anova",
  control = myc
)
plot(bos.to, compress = TRUE) # type='proportional')
```

Not surprisingly, the predictions of this large tree are 
not very good:
```{r prune2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
# predictions are poor, unsurprisingly
pr.to <- predict(bos.to, newdata = dat.te, type = "vector")
with(dat.te, mean((medv - pr.to)^2))
```

To prune we explore the *CP table* returned in the
`rpart` object to find the value of the complexity
parameter with optimal estimated prediction error. The estimated
prediction error of each subtree (corresponding to each value of `CP`)
is contained in the column `xerror`, and the associated 
standard deviation is in column `xstd`. We would like to find
the value of `CP` that yields a corresponding pruned tree with smallest
estimated prediction error. The function `printcp` shows the
CP table corresponding to an `rpart` object:
```{r prune3, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
printcp(bos.to)
```
It is probably better and easier to find this 
optimal value *programatically* as follows: 
```{r prune4.2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
(b <- bos.to$cptable[which.min(bos.to$cptable[, "xerror"]), "CP"])
```
<!-- > **R coding digression**: Note that above we could also have used the following: -->
<!-- > ```{r prune4.alt, fig.width=6, fig.height=6, message=FALSE, warning=FALSE} -->
<!-- > tmp <- bos.to$cptable[,"xerror"] -->
<!-- > (b <- bos.to$cptable[ max( which(tmp == min(tmp)) ), "CP"] ) -->
<!-- > ``` -->
<!-- > What is the difference between `which.min(a)` and `max( which( a == min(a) ) )`? -->

We can now use the function 
`prune` on the `rpart` object setting the complexity parameter
to the estimated optimal value found above:
```{r prune4, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
bos.t3 <- prune(bos.to, cp = b)
```

This is how the optimally pruned tree looks:
```{r prune4.5, fig.width=8, fig.height=8, message=FALSE, warning=FALSE}
plot(bos.t3, uniform = FALSE, margin = 0.01)
text(bos.t3, pretty = FALSE)
```

Finally, we can check the predictions of the pruned 
tree on the test set:
```{r prune5, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
# predictions are better
pr.t3 <- predict(bos.t3, newdata = dat.te, type = "vector")
with(dat.te, mean((medv - pr.t3)^2))
```

Again, it would be a **very good exercise** for you to 
compare the MSPE of the pruned tree with that of several
of the alternative methods we have seen in class so far,
**without using a training / test split**. 



## Pruning regression trees with `tree`

The implementation of trees in the `R` package `tree` follows
the original CV-based pruning strategy, as discussed in 
Section 3.4 of the book

> Breiman, L., Friedman, J.H., Olshen, R.A. and Stone, C.J. (1984). Classification and regression trees. Chapman & Hall.

or Section 7.2 of:

> Ripley, Brian D. (1996). Pattern recognition and neural networks. Cambridge University Press

Both books are available in electronic form from the UBC Library:
[Breiman et al.](http://tinyurl.com/y3g2femt) and 
[Ripley, B.D.](http://tinyurl.com/yylchlys).

We now use the function `tree::tree()` to fit the same regression
tree as above. Note that the default stopping criteria in this
implementation of regression trees is different from the one in
`rpart::rpart()`, hence to obtain the same results as above we
need to modify the default stopping criteria using the argument
`control`: 
```{r prunetree0, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(tree)
bos.t2 <- tree(medv ~ ., data = dat.tr, control = tree.control(nobs = nrow(dat.tr), mincut = 6, minsize = 20))
```

We plot the resulting tree

```{r prunetree1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(bos.t2)
text(bos.t2)
```

As discussed before, we now fit a very large tree, which will be 
pruned later:

```{r prunetree2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, pretty=TRUE}
set.seed(123)
bos.to2 <- tree(medv ~ .,
  data = dat.tr,
  control = tree.control(nobs = nrow(dat.tr), mincut = 1, minsize = 2, mindev = 1e-5)
)
plot(bos.to2)
```

We now use the function `tree:cv.tree()` to estimate the MSPE of 
the subtrees of `bos.to2`, using 5-fold CV, and plot the estimated
MSPE (here labeled as "deviance") as a function of the 
complexity parameter (or, equivalently, the size of the tree): 
```{r prunetree3, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
set.seed(123)
tt <- cv.tree(bos.to2, K = 5)
plot(tt)
```

Finally, we use the function `prune.tree` to prune the larger tree
at the "optimal" size, as estimated by `cv.tree` above:
```{r prunetree3.2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
bos.pr2 <- prune.tree(bos.to2, k = tt$k[max(which(tt$dev == min(tt$dev)))])
plot(bos.pr2)
text(bos.pr2)
```

Compare this pruned tree with the one obtained with the regression trees
implementation in `rpart`. In particular, we can compare the
predictions of this other pruned 
tree on the test set:
```{r prune.tree.pred, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
# predictions are worse than the rpart-pruned tree
pr.tree <- predict(bos.pr2, newdata = dat.te, type = "vector")
with(dat.te, mean((medv - pr.tree)^2))
```
Note that the predictions of the tree pruned with the `tree` 
package seem to be better than those of the tree pruned with 
the `rpart` package. **Does this mean that `rpart` gives
trees with worse predictions than `tree` for data coming
from the process than generated our training set?** 
**Or could it all be an artifact of the specific test set we used?**
**Can you think of an experiment to check this?**
Again, it would be a **very good exercise** for you to 
check which fit (`tree` or `rpart`) gives pruned 
trees with better prediction properties in this case. 


## Instability of regression trees 

Trees can be rather unstable, in the sense that small changes in the
training data set may result in relatively large differences in the
fitted trees. As a simple illustration we randomly split the 
`Boston` data used before into two halves and fit a regression
tree to each portion. We then display both trees.

```{r inst1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
# Instability of trees...
library(rpart)
data(Boston, package = "MASS")
set.seed(123)
n <- nrow(Boston)
ii <- sample(n, floor(n / 2))
dat.t1 <- Boston[-ii, ]
bos.t1 <- rpart(medv ~ ., data = dat.t1, method = "anova")
plot(bos.t1, uniform = FALSE, margin = 0.01)
text(bos.t1, pretty = TRUE, cex = .8)
```

```{r inst2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
dat.t2 <- Boston[ii, ]
bos.t2 <- rpart(medv ~ ., data = dat.t2, method = "anova")
plot(bos.t2, uniform = FALSE, margin = 0.01)
text(bos.t2, pretty = TRUE, cex = .8)
```

Although we would expect both random halves of the same (moderately large) 
training set to beat least qualitatively similar, 
Note that the two trees are rather different. 
To compare with a more stable predictor, we fit a linear
regression model to each half, and look at the two sets of estimated
coefficients side by side:
```{r inst3, fig.width=5, fig.height=5, message=FALSE, warning=FALSE}
# bos.lmf <- lm(medv ~ ., data=Boston)
bos.lm1 <- lm(medv ~ ., data = dat.t1)
bos.lm2 <- lm(medv ~ ., data = dat.t2)
cbind(
  round(coef(bos.lm1), 2),
  round(coef(bos.lm2), 2)
)
```
Note that most of the estimated regression coefficients are 
similar, and all of them are at least qualitatively comparable. 

