# What is Adaboost doing, *really*? 

Following the work of [@FriedmanHastie2000] (see also
Chapter 10 of [ESL]), we saw in class that Adaboost can be 
interpreted as fitting an *additive model* in a stepwise (greedy) way,
using an exponential loss. 
It is then easy to prove that Adaboost.M1 
is computing an approximation to the *optimal classifier*
G( x ) = log[ P( Y = 1 | X = x ) / P( Y = -1 | X = x ) ] / 2,
where *optimal* here is taken with respect to the **exponential loss** 
function. More specifically, Adaboost.M1 is using an 
additive model to approximate that function. In other words, Boosting is
attempting to find functions $f_1$, $f_2$, ..., $f_N$ such that 
$G(x) = \sum_i f_i( x^{(i)} )$, where $x^{(i)}$ is a sub-vector
of $x$ (i.e. the function $f_i$ only depends on *some* of the
available features, typically a few of them: 1 or 2, say). Note
that each $f_i$ generally depends on a different subset of 
features than the other $f_j$'s. 

Knowing the function the boosting algorithm is approximating (even
if it does it in a greedy and suboptimal way), allows us to 
understand when the algorithm is expected to work well,
and also when it may not work well. 
In particular, it provides one way to choose the complexity of the 
*weak lerners* used to construct the ensemble. For an example
you can refer to the corresponding lab activity. 

### A more challenging example, the `email spam` data

The email spam data set is a relatively classic data set 
containing 57 features (potentially explanatory variables) 
measured on 4601 email messages. The goal is to predict
whether an email is *spam* or not. The 57 features are 
a mix of continuous and discrete variables. More information
can be found at
[https://archive.ics.uci.edu/ml/datasets/spambase](https://archive.ics.uci.edu/ml/datasets/spambase).

We first load the data and randomly separate it into a training and
a test set. A more thorough analysis would be to use 
*full* K-fold cross-validation, but given the computational
complexity, I decided to leave the rest of this 
3-fold CV exercise to the reader. 
```{r spam.1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
data(spam, package = "ElemStatLearn")
n <- nrow(spam)
set.seed(987)
ii <- sample(n, floor(n / 3))
spam.te <- spam[ii, ]
spam.tr <- spam[-ii, ]
```
We now use Adaboost with 500 iterations, using *stumps* (1-split
trees) as our
weak learners / classifiers, and check the performance on
the test set:
```{r spam.2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(adabag)
onesplit <- rpart.control(cp = -1, maxdepth = 1, minsplit = 0, xval = 0)
bo1 <- boosting(spam ~ ., data = spam.tr, boos = FALSE, mfinal = 500, control = onesplit)
pr1 <- predict(bo1, newdata = spam.te)
table(spam.te$spam, pr1$class) # (pr1$confusion)
```
The classification error rate on the test set is `r round(pr1$error, 3)`. We now
compare it with that of a Random Forest and look at the fit:
```{r spam.3, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(randomForest)
set.seed(123)
(a <- randomForest(spam ~ ., data = spam.tr, ntree = 500))
```
Note that the OOB estimate of the classification error rate 
is `r round(a$err.rate[a$ntree, "OOB"], 3)`. 
The number of trees used seems to be appropriate in terms
of the stability of the OOB error rate estimate:
```{r spam.plot.rf, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(a)
```

Now use the test set to estimate the error rate of the Random Forest 
(for a fair comparison with the one computed with boosting) and obtain
```{r spam.error.rate.rf, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
pr.rf <- predict(a, newdata = spam.te, type = "response")
table(spam.te$spam, pr.rf)
```
The performance of Random Forests on this test set is better than that of 
boosting (recall that the estimated classification error rate 
for 1-split trees-based Adaboost was 
`r round(pr1$error, 3)`, while for the Random Forest is `r round(mean( spam.te$spam != pr.rf ), 3)` on the test set and `r round(a$err.rate[a$ntree, "OOB"], 3)` using OOB). 

Is there *any room for improvement* for Adaboost? 
As we discussed in class, depending on the interactions that may be 
present in the *true classification function*, we might be able to 
improve our boosting classifier by slightly increasing the complexity
of our base ensemble members. Here we try to use 3-split classification
trees, instead of the 1-split ones used above:
```{r spam.4, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
threesplits <- rpart.control(cp = -1, maxdepth = 3, minsplit = 0, xval = 0)
bo3 <- boosting(spam ~ ., data = spam.tr, boos = FALSE, mfinal = 500, control = threesplits)
pr3 <- predict(bo3, newdata = spam.te)
(pr3$confusion)
```
The number of elements on the boosting ensemble (500) appears to be
appropriate when we look at the error rate on the test set as 
a function of the number of boosting iterations:
```{r spam.5, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(errorevol(bo3, newdata = spam.te))
```

There is, in fact, a noticeable improvement in performance on this
test set compared to the AdaBoost using *stumps*. 
The estimated classification error rate of AdaBoost using 3-split trees on this test set is
`r round(pr3$error,3)`. Recall that the estimated classification error rate
for the Random Forest was `r round(mean( spam.te$spam != pr.rf ),3)`
(or `r round(a$err.rate[a$ntree, "OOB"],3)` using OOB). 

As mentioned above you are strongly encouraged to finish this analysis
by doing a complete K-fold CV analysis in order to compare boosting with random 
forests on these data. 


### An example on improving Adaboost's performance including interactions

**Some error I can't track happens below**

Consider the data set in the file `boost.sim.csv`. This 
is a synthetic data inspired by the 
well-known Boston Housing data. The response variable is `class` 
and the two predictors are `lon` and `lat`. We read the data set
```{r sim.read}
sim <- read.table("data/boost.sim.csv", header = TRUE, sep = ",", row.names = 1)
```
We split the data randomly into a training and a test set:
```{r sim.split}
set.seed(123)
ii <- sample(nrow(sim), nrow(sim) / 3)
sim.tr <- sim[-ii, ]
sim.te <- sim[ii, ]
```
As before, we use *stumps* as our base classifiers
```{r boston0}
library(rpart)
stump <- rpart.control(cp = -1, maxdepth = 1, minsplit = 0, xval = 0)
```
and run 300 iterations of the boosting algorithm:
```{r bostonsimu0, eval=FALSE}
set.seed(17)
sim1 <- boosting(class ~ ., data = sim.tr, boos = FALSE, mfinal = 300, control = stump)
```
We examine the evolution of our ensemble on the test set:
```{r bostonsimu1, eval=FALSE}
plot(errorevol(sim1, newdata = sim.te))
```

and note that the peformance is both disappointing and does not improve with
the number of iterations. The error rate on the test set is
`r print("uhoh")`.
Based on the discussion in class about the effect of the 
complexity of the base classifiers, 
we now increase slightly their complexity: from
stumps to trees with up to 2 splits:
```{r boston00, eval=FALSE}
twosplit <- rpart.control(cp = -1, maxdepth = 2, minsplit = 0, xval = 0)
set.seed(17)
sim2 <- boosting(class ~ ., data = sim.tr, boos = FALSE, mfinal = 300, control = twosplit)
plot(errorevol(sim2, newdata = sim.te))
```

Note that the error rate improves noticeably to 
`r print("less uhoh")`.
Interestingly, note as well that increasing the number 
of splits of the base classifiers does not seem to 
help much. With 3-split trees:
```{r boston00.more, eval=FALSE}
threesplit <- rpart.control(cp = -1, maxdepth = 3, minsplit = 0, xval = 0)
set.seed(17)
sim3 <- boosting(class ~ ., data = sim.tr, boos = FALSE, mfinal = 300, control = threesplit)
plot(errorevol(sim3, newdata = sim.te))
```
```{r boston001, echo=TRUE, eval=FALSE}
foursplit <- rpart.control(cp = -1, maxdepth = 4, minsplit = 0, xval = 0)
set.seed(17)
sim4 <- boosting(class ~ ., data = sim.tr, boos = FALSE, mfinal = 300, control = foursplit)
```
the error rate on the test set is 
```{r, eval=FALSE}
round(predict(sim3, newdata = sim.te)$error, 4)
```
while with 4-split trees the error rate is 
```{r, eval=FALSE}
round(predict(sim4, newdata = sim.te)$error, 4)
```


The explanation for this is that the response variables 
in the data set \texttt{boost.sim.csv} were in fact generated
through the following relationship:
```
log [ P ( Y = 1 | X = x ) / P ( Y = -1 | X = x ) ] / 2
 = [ max( x2 - 2, 0) - max( x1 + 1, 0) ] ( 1- x1 + x2 )
```
where $x = (x_1, x_2)^\top$. Since *stumps* (1-split trees) 
are by definition functions of a single
variable, boosting will not be able to approximate the above function using
a linear combination of them, regardless of how many terms you use. Two-split
trees, on the other hand, are able to model interactions between the two
explanatory variables $X_1$ (`lon`) and
$X_2$ (`lat`), and thus, with sufficient terms in the sum, we are able to 
approximate the above function relatively well. 

As before, note that the analysis above may depend on the specific 
training / test split we used, so it is strongly suggested that you
re-do it using a proper cross-validation setup. 


